---
title: "Comparing GWB thermodynamic data files"
output:
  html_vignette:
    mathjax: null
vignette: >
  %\VignetteIndexEntry{1 - Comparing two GWB thermodynamic data files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: logKcalc.bib
csl: elementa.csl
---

<!-- https://gomakethings.com/how-to-break-an-image-out-of-its-parent-container-with-css/ -->
<style>
@media (min-width: 700px) {
  .full-width {
    left: 50%;
    margin-left: -50vw;
    margin-right: -50vw;
    max-width: 100vw;
    position: relative;
    right: 50%;
    width: 100vw;
  }
}
@media (min-width: 1020px) {
  .full-width {
    left: 50vw; /* fallback if needed */
    left: calc(50vw - 160px);
    width: 1000px;
    position: relative;
    background-color: #9ecff7;
    padding:10px;
  }
}
</style>

```{r setup, include=FALSE}
library(logKcalc)
library(knitr)
## colorize messages 20171031
## adapted from https://gist.github.com/yihui/2629886#file-knitr-color-msg-rnw
color_block = function(color) {
  function(x, options) sprintf('<pre style="color:%s">%s</pre>', color, x)
}
knit_hooks$set(warning = color_block('magenta'), error = color_block('red'), message = color_block('blue'))
## use pngquant to optimize PNG images
knit_hooks$set(pngquant = hook_pngquant)
pngquant <- "--speed=1 --quality=0-25"
if (!nzchar(Sys.which("pngquant"))) pngquant <- NULL 
```

This vignette was compiled with **logKcalc** `r packageDescription("logKcalc")$Version` and **CHNOSZ** `r packageDescription("CHNOSZ")$Version`.

#### These plots compare the logK values in thermodynamic data files from GWB and K2GWB.

The file `thermo_24noredox.tdat` was derived from the `thermo.tdat` file distributed with The Geochemist's Workbench® to write all dissociation reactions in terms of the basis species, not with redox couples.
The file `ThermoGWB_15_6_2020.tdat` was created in K2GWB [@CB05] using the same basis species.
Both files have the same 24 elements (Ag, Al, As, Au, C, Ca, Cl, Cu, F, Fe, H, K, Mg, Mn, N, Na, O, P, Pb, S, Si, Sn, U, Zn).

The logKcomp function reads the reaction coefficients and log*K* values from the files and only considers species with the same dissociation reactions in both files.
The plot titles indicate how many of these species have log*K* values that are not available (NA) at this particular temperature and pressure (indicated by a "500" in the GWB file).
The function adds labels to points with a difference of greater that 1 log*K* unit.

One type of species block can be selected for comparison (aqueous, mineral, gas); here we choose minerals.
The coordinates are manually adjusted to exclude some points at extreme values, but no labels are lost; the line segments point to the points off the scale.

```{r logKcomp1-1, fig.width = 12, fig.height = 6, out.width = "100%", out.extra='class="full-width"', pngquant = pngquant}
file1 <- system.file("extdata/ThermoGWB_15_6_2020.tdat", package = "logKcalc")
file2 <- system.file("extdata/thermo_24noredox.tdat", package = "logKcalc")
lab1 <- "UNITHERM"
lab2 <- "thermo.tdat"
plot1 <- logKcomp(file1, file2, "mineral", 2, lab1, lab2, c(-25, 200), c(-5, 10))
```

The first plot is for 2nd *T*,*P* pair (25 °C); now we make one for the 8th (300 °C).
This time the messages about unmatched species between the files are hidden.

```{r logKcomp1-2, message = FALSE, results = "hide", fig.width = 12, fig.height = 6, out.width = "100%", out.extra='class="full-width"', pngquant = pngquant}
plot2 <- logKcomp(file1, file2, "mineral", 8, lab1, lab2, c(-25, 100))
gridExtra::grid.arrange(plot1, plot2, ncol = 2)
```

The plots reveal systematic differences between the databases: most of the large differences correspond to higher log*K* values for the mineral dissociation reactions in `thermo.tdat`, which indicates overall lower stability of the minerals.
These differences may be associated with different estimation algorithms for thermodynamic parameters (e.g. extrapolation from high-temperature phase equilibrium data or derivation from primarily calorimetric measurements).

#### Here are some recommended commands for processing the GWB data file created by K2GWB to recalculate the log*K* values using the [OBIGT database in CHNOSZ](../../CHNOSZ/doc/obigt.html).

The `modOBIGT` command adds more minerals from the SUPCRT92 database, an entry for steam that has been removed from OBIGT, and a modified version of realgar to match the chemical formula used in UNITHERM.

```{r logKcalc}
infile <- system.file("extdata/ThermoGWB_15_6_2020.tdat", package = "logKcalc")
outfile <- file.path(tempdir(), "ThermoGWB_OBIGT.tdat")
reset()
modOBIGT(c("addSUPCRT", "steam", "realgar*4"))
logKcalc(infile, outfile)
```

#### These plots compare the logK values in the K2GWB file and OBIGT.

Only species with the same name and dissociation reaction in both files are considered for the plot.
Labels are added to points with a difference of greater that 1 log*K* unit.

```{r logKcomp2, message = FALSE, results = "hide", fig.width = 12, fig.height = 6, out.width = "100%", out.extra='class="full-width"', pngquant = pngquant}
lab2 <- "OBIGT"
plot1 <- logKcomp(infile, outfile, "mineral", 2, lab1, lab2)
plot2 <- logKcomp(infile, outfile, "mineral", 8, lab1, lab2)
gridExtra::grid.arrange(plot1, plot2, ncol = 2)
```

There are again some differences, but the mineral data in UNITHERM and OBIGT are on average more similar to each other than to `thermo.tdat`.

## References
