---
title: "Processing a GWB thermodynamic data file"
output:
  html_vignette:
    mathjax: null
vignette: >
  %\VignetteIndexEntry{2 - Processing a GWB thermodyamic data file}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: logKcalc.bib
csl: elementa.csl
---

<!-- https://gomakethings.com/how-to-break-an-image-out-of-its-parent-container-with-css/ -->
<style>
@media (min-width: 700px) {
  .full-width {
    left: 50%;
    margin-left: -50vw;
    margin-right: -50vw;
    max-width: 100vw;
    position: relative;
    right: 50%;
    width: 100vw;
  }
}
@media (min-width: 1020px) {
  .full-width {
    left: 50vw; /* fallback if needed */
    left: calc(50vw - 160px);
    width: 1000px;
    position: relative;
    background-color: #9ecff7;
    padding:10px;
  }
}
</style>

```{r setup, include=FALSE}
library(logKcalc)
library(knitr)
## colorize messages 20171031
## adapted from https://gist.github.com/yihui/2629886#file-knitr-color-msg-rnw
color_block = function(color) {
  function(x, options) sprintf('<pre style="color:%s">%s</pre>', color, x)
}
knit_hooks$set(warning = color_block('magenta'), error = color_block('red'), message = color_block('blue'))
## use pngquant to optimize PNG images
knit_hooks$set(pngquant = hook_pngquant)
pngquant <- "--speed=1 --quality=0-25"
if (!nzchar(Sys.which("pngquant"))) pngquant <- NULL 
```

#### This vignette shows the results from processing thermo_24elements.tdat.

This vignette was compiled with **logKcalc** `r packageDescription("logKcalc")$Version` and **CHNOSZ** `r packageDescription("CHNOSZ")$Version`.

The thermodynamic system is defined by `thermo_24elements.tdat`, which is a modfied version of GWB's `thermo.tdat` file reduced to 24 elements.
First we use `modOBIGT` to add some minerals from SUPCRT92, an entry for steam that has been removed from OBIGT, and antigorite with a chemical formula to match the one used in thermo.tdat.

```{r modOBIGT, message = FALSE}
reset()
modOBIGT(c("addSUPCRT", "steam", "antigorite/2"))
```

Some basis species used in the GWB file are not in the default OBIGT database.
They can be added to OBIGT by fitting their thermodynamic parameters (Δ*G*°, *S*° and *C~p~*°) to the log*K* values for their dissociation reactions from a different GWB file where they have been swapped out of the basis.

```{r addOBIGT, message = FALSE}
addOBIGT("As(OH)4-")
addOBIGT("Sn++++")
```

Here we set the file paths and perform the log*K* calculation.
The `maxprint` argument is used to prevent the printing of long lists of species that can't be found in OBIGT.

```{r logKcalc}
infile <- system.file("extdata/thermo_24elements.tdat", package = "logKcalc")
outfile <- file.path(tempdir(), "thermo_OBIGT.tdat")
logKcalc(infile, outfile, maxprint = 10)
```

Note that species are removed if their log*K* values are all NA.
This includes a few minerals taken from SUPCRT92, as well as pyrrhotite, whose chemical formula in the GWB file (Fe.875S) is different from that in OBIGT (FeS).

#### Now we can make some plots to compare the logK values in the input and output files.

Only species with the same name and dissociation reaction in both files are considered for the plot.
The plot titles indicate how many of these species have log*K* values that are not available (NA) at this particular temperature and pressure (indicated by a "500" in the GWB file).
Labels are added to points with a difference of greater that 1 log*K* unit.

```{r logKcomp, message = FALSE, results = "hide", fig.width = 12, fig.height = 6, out.width = "100%", out.extra='class="full-width"', pngquant = pngquant}
plot1 <- logKcomp(infile, outfile, iTP = 4, lab2 = "OBIGT")
plot2 <- logKcomp(infile, outfile, type = "mineral", iTP = 4, lab2 = "OBIGT")
gridExtra::grid.arrange(plot1, plot2, ncol = 2)
```
