---
title: "Using fitted thermodynamic parameters in CHNOSZ"
output:
  html_vignette:
    mathjax: null
vignette: >
  %\VignetteIndexEntry{5 - Using fitted thermodynamic parameters in CHNOSZ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: logKcalc.bib
csl: elementa.csl
---

<!-- https://gomakethings.com/how-to-break-an-image-out-of-its-parent-container-with-css/ -->
<style>
@media (min-width: 700px) {
  .full-width {
    left: 50%;
    margin-left: -50vw;
    margin-right: -50vw;
    max-width: 100vw;
    position: relative;
    right: 50%;
    width: 100vw;
  }
}
@media (min-width: 1020px) {
  .full-width {
    left: 50vw; /* fallback if needed */
    left: calc(50vw - 160px);
    width: 1000px;
    position: relative;
    background-color: #9ecff7;
    padding:10px;
  }
}
</style>

```{r setup, include=FALSE}
library(logKcalc)
library(knitr)
## colorize messages 20171031
## adapted from https://gist.github.com/yihui/2629886#file-knitr-color-msg-rnw
color_block = function(color) {
  function(x, options) sprintf('<pre style="color:%s">%s</pre>', color, x)
}
knit_hooks$set(warning = color_block('magenta'), error = color_block('red'), message = color_block('blue'))
## use pngquant to optimize PNG images
knit_hooks$set(pngquant = hook_pngquant)
pngquant <- "--speed=1 --quality=0-25"
if (!nzchar(Sys.which("pngquant"))) pngquant <- NULL 
## logK with a thin space 20200627
logK <- "log&thinsp;<i>K</i>"
## run reset() here so we don't use the DEW model (from vig4.Rmd)
reset()
```

#### This vignette shows how to use data for species from a GWB file in a diagram made with CHNOSZ.

This vignette was compiled with **logKcalc** `r packageDescription("logKcalc")$Version` and **CHNOSZ** `r packageDescription("CHNOSZ")$Version`.

One of the example scripts for the Act2 program in The Geochemist's Workbench® is a mosaic diagram for the As-S-O-H system [@GWB20a].
A similar diagram can be produced in CHNOSZ, but some differences are apparent because of different thermodynamic data.
The source of thermodynamic data for As-bearing species in this diagram is @NA03.
See the [As_mosaic.R](As_mosaic.R) file in this directory for the script used to make this diagram.

```{r Mosaic1, message = FALSE, fig.width = 12, fig.height = 10, out.width = "75%", fig.align = "center", pngquant = pngquant}
source("As_mosaic.R")
```

In order to more closely reproduce the diagram from GWB, we can use `addOBIGT()` to fit thermodynamic parameters to the `r logK` values of dissociation reactions of species from a GWB data file.
This function uses a specially modified version of GWB's `thermo.tdat` in which the basis species are all available in OBIGT (`thermo_24swapped.tdat`).
The fitted parameters (Δ*G*°, *S*° and *C~p~*°) of the formed species are added to the OBIGT database, so they can be used in diagrams made in CHNOSZ.

Here, we fit the parameters for minerals and aqueous species other than As(OH)~3~ (H~3~AsO~3~ in OBIGT), which is a basis species in `thermo_24swapped.tdat`.
Although it has no stability field in the diagram, AsH~3~ is also updated because it appears in the dissociation reaction for realgar.
Because it uses just three parameters, the fit may introduce some error, so we increase the tolerance (in `r logK` units) for H~2~AsO~4~^-^ to allow the checks for this species to pass.
We run `reset()` first to make sure we are starting with the default OBIGT database.

```{r addOBIGT, message = FALSE}
reset()
addOBIGT("AsH3(aq)")
addOBIGT("H2AsO4-", tolerance = 0.15)
addOBIGT("HAsO4--")
addOBIGT("As(OH)4-")
addOBIGT("AsO2OH--")
addOBIGT("AsO4---")
addOBIGT("Orpiment", "As2S3")
addOBIGT("Realgar", "AsS")
```

The warning about an unbalanced reaction occurs because of the inexact representation of reaction coefficients in the GWB file due to rounding, and can be ignored.

For completeness we can even use fitted thermodynamic parameters for the S-bearing basis species that are speciated across the mosaic diagram.
This does not include SO~4~^-2^, since it is a basis species in `thermo_24swapped.tdat`.
For the purposes of the reproduction, we include S^-2^, which has been recommended to be removed from thermodynamic databases [@MBH+18] and is not present in the default OBIGT database.

```{r addS, message = FALSE}
addOBIGT("HSO4-")
addOBIGT("H2S(aq)")
addOBIGT("HS-")
addOBIGT("S--")
```

Note that repeated '-' or '+' signs at the end of formulas are converted into e.g. '-2' for compatibility with the OBIGT database.

Next we identify As-bearing species in the As-S-O-H system.
We remove species with data from @NA03, as they may conflict with the GWB dataset.
However, we explicitly include H~3~AsO~3~ here; although its thermodynamic parameters in OBIGT are still from @NA03, the parameters of the *other* species were made to be consistent with the `r logK` values for the dissociation reactions in `thermo_24swapped.tdat` that involve H~3~AsO~3~ as a basis species.

```{r species, message = FALSE}
iaq <- retrieve("As", c("S", "O", "H"), "aq")
iaq <- iaq[!grepl("NA03", info(iaq)$ref1)]
iaq <- c(info("H3AsO3"), iaq)
icr <- retrieve("As", c("S", "O", "H"), "cr")
icr <- icr[!grepl("NA03", info(icr)$ref1)]
```

Now we write a function to set up the system in CHNOSZ and make the mosaic diagram.
The activity of As-bearing aqueous species is set to 10^-3^ and that of SO~4~^-2^ to 10^-4^ (this value applies to the S-bearing basis species as a whole).

```{r mosaicfun}
mosaicfun <- function(blend = FALSE) {

  # Set up basis species and system
  basis(c("H3AsO3", "SO4-2", "H2O", "O2", "H+"))
  basis("SO4-2", -4)
  species(iaq, -3)
  species(icr, 0)
  # Set pH and loga(O2) range and grid resolution
  pH <- c(0, 14, 500)
  O2 <- c(-70, -40, 500)
  # Calculate reaction affinities for changing basis species
  bases <- c("SO4-2", "HSO4-", "H2S", "HS-", "S-2")
  m <- mosaic(bases, pH = pH, O2 = O2, T = 100, blend = blend)
  # Make the diagram
  par(cex = 2.5, mar = c(2.8, 3, 1.4, 1))
  fill <- c(rep("aliceblue", length(iaq)), rep("antiquewhite1", length(icr)))
  d <- diagram(m$A.species, fill = fill, lwd = 2)
  diagram(m$A.bases, add = TRUE, col = "blue", lty = 3, lwd = 2,
          col.names = "blue", cex.names = 0.7, italic = TRUE)
  water.lines(d)
  legend("bottomleft", describe.property("T", 100), bty = "n")
  title(main = paste("blend =", blend), font.main = 1)

}
```

Here are two diagrams, the first made with blend = FALSE to portray sharp boundaries between basis species, and the second made with blend = TRUE to represent the continuous change of speciation between basis species.

```{r Mosaic2, fig.width = 24, fig.height = 12, out.width = "100%", out.extra='class="full-width"', message = FALSE, pngquant = pngquant}
par(mfrow = c(1, 2))
mosaicfun(FALSE)
mosaicfun(TRUE)
```

The diagram on the left is very similar to the one from the GWB example [@GWB20a `p.98`].

## References
